### 2022-09-30 main/chapter_searches.tex
overleaf readthrough

\begin{singlespacing}
\chapter{Data analysis and \atlas\ searches}
\label{chapter:searches}
\begin{epigraphs}
\qitem{%
One way is to make it so simple that there are \emph{obviously} no deficiencies
and the\\
other way is to make it so complicated that there are no \emph{obvious}
deficiencies.%
}%
{Tony~Hoare,
\textit{The Emperor’s Old Clothes},
1981~\cite{hoare2007emperor}}
\end{epigraphs}
\end{singlespacing}
\noindent
Our aim in scientific data analysis is to distil useful messages from the
data, and to report that message to the wider community.
Various messages could be extracted from given data,
and those messages will have differing values to their recipients.
Data analysis is not an automatic process, but a design with many judgements
that are informed by theory, intuition, and convention to give value to their
results.

To support our later description of the \atlas\ SUSY $\twoljets$ search in
Chapter~\ref{chapter:2ljets}, this chapter introduces the basic theory and
practice of data analysis and its conventional form in the \atlas\ SUSY
community, parts of which can only be explained in their historical context.

@@

\section{Data analysis}
\label{sec:searches_data_analysis}
How many jelly beans are in the jar of Figure~\ref{fig:searches_beans}?
Your answer depends not only on what you observe from the figure, but also on
what else you know about cartoon jars of beans, and on what reward is offered.
Each answer is a decision, which depends both on how you value the rewards,
and on which numbers of beans you subjectively infer to be probable.
We shall return to this example after developing its analysis in probability
and decision theory.

@@

Probabilistic data analysis follows the ordinary structure of science:
write down predictive models, test those predictions against data,
then update your models (or invent new ones) to refine their predictions
to accommodate what has been learned.
Likelihood functions are particularly important to this process because they
describe our predictions.
But as we shall see, not everything is a likelihood.

We begin with what probability is: at its most abstract, a probability is a
proportion of stuff within a larger whole.
Probabilities are useful because proportions have applications
(particularly to logical reasoning with uncertainty), and because
they can be manipulated and interrelated with an intuitive linear algebra.
The remainder of this section uses a standard example from High Energy Physics
--- a histogram ---
to introduce this probabilistic structure and other parts of its technical
language.

\subsection{Probability: abstract}
A probability $\prob{b}{a}$ is a number, which quantifies the proportion
of $b$ within $a$ and can be expressed as a ratio~\cite{axioms1010038}
\begin{equation}
\label{eqn:searches_prob_ratio}
\prob{b}{a} = \frac{m(b, a)}{m(a)},
\end{equation}
in which $m(x)$ is a measure, and the comma `,' means `and'
(logical and or set intersection or anything equivalent).

@@

Logic is an important application of probability.
To apply to logical reasoning, $a$ and $b$ are logical propositions and
$\prob{b}{a}$ quantifies the degree to which $a$ implies $b$, with the
limiting cases that $\prob{b}{a} = 1$ when $a$ implies $b$,
and $\prob{b}{a} = 0$ when $a$ contradicts $b$.
Indeed, Cox's theorem shows that probability algebra is uniquely consistent
with Boolean logic~\cite{
cox1946probability,
cox1961algebra,
garrett1998nand,
jaynes2003probability,
keynes1920treatise
}.
Probability works for logic, and it works \emph{for tea, too}.

@@

Standard probability algebra derives from
Equation~\ref{eqn:searches_prob_ratio}; it comprises a \textbf{product rule}
\begin{align}
\prob{c, b}{a} &= \prob{c}{b, a}\times \prob{b}{a},
\label{eqn:searches_product_rule}
\intertext{%
for nested proportions
(a spoon in a teacup in a sink),
and a \textbf{sum rule}%
}
\prob{c\;\textrm{or}\;b}{a} &= \prob{c}{a} + \prob{b}{a} - \prob{c, b}{a},
\label{eqn:searches_sum_rule}
\end{align}
for combinations in a union, with sums over exhaustive sets of outcomes
necessarily normalized to unity.

@@

Since prior probabilities are normalized distributions, it is usually
convenient to represent them as density functions.
Densities are stated relative to coordinates (or occasionally measures),
so changing coordinates changes their values as their unit probability mass
is sloshed around.
Likelihoods are not densities in the conditioning variable that they explore,
so changing coordinates instead slides them horizontally like beads on an
abacus, and does not change their values for equivalent states.
Figure~\ref{fig:searches_abacus} illustrates this difference.

@@

To illustrate this, consider a histogram that bins event yields in
$\met$, with distributions from background and (supersymmetric) signal samples
overlaid.
Example such histograms are displayed in
Figure~\ref{fig:searches_sig_bkg_prior_likelihood}.
Since these histograms are normalized to unit area, they show prior
probability distributions along the bin axis.
There is one prior for events from each of the background and signal
hypotheses ---
priors distribute unit probability explore along the bin axis.
In each bin, however, a likelihood function compares the assignments from
the two hypotheses: signal and background ---
likelihoods explore vertically down the sample axis.

@@

Both distributions are equivalently drawn in two different coordinates
on the left and right of Figure~\ref{fig:searches_sig_bkg_prior_likelihood};
they are identical in content, but differ in appearance because they represent
each prior distribution as a density in the standard way.%
\footnote{%
Density is a somewhat misleading term here ---
probability mass is incompressible; it moves and redistributes more like
depth of water than density of air.%
}

To preserve the area of each bin, which shows its probability mass,
bin heights therefore change to adapt to their widths that change due to the
differing values of $\met$ and $\sqrt{\met}$.
This coordinate-dependence of the prior density means that its point of
maximum density is not always meaningful.
But since the probability mass in each bin is constant, the likelihood is
coordinate-independent.

Standard methods approximate such histograms from Monte Carlo simulations,
which do not directly indicate bin indices but state the values for event
variables.
To connect event variables to bin indices, we begin with some notation:
we have an hypothesis label $h$ (signal or background),
an event variable $x$ that we histogram ($x = \met$),
and we observe a datum $d$ that indexes the histogram bins.
Supposing, for example, that an event is observed in the overflow bin, it says
$d = `x > 100\,\eV[G]\textrm{'}$.
This observed bin has a likelihood function
\begin{equation}
L(x) = \prob{d}{x, h} = \prob{d}{x} =
\left\{
\begin{matrix}
1 & \textrm{if }~x > 100\,\eV[G], \\
0 & \textrm{otherwise} \\
\end{matrix}
\right.
,
\end{equation}
which corresponds to Equation~\ref{eqn:searches_likelihood} by
$b \leftarrow d$ and $a \leftarrow `x, h\textrm{'}$,
and simplifies by $\prob{d}{x, h} = \prob{d}{x}$ because at given $x$ the
binning is independent of any origin from signal or background.%
\footnote{%
Independence of the likelihood function from prior hypotheses is so common
that it should usually be assumed unless otherwise stated.
}
This example of a binary selection highlights how the likelihood acts as a
filter that only rejects events that are contradicted by the data.
Other likelihoods that are not binary act similarly, but as inefficient filters
that reject events with some probability.

@@

When we fill a histogram by summing weights of simulated samples, we are
calculating another likelihood over hypotheses $h$ alone in the bin,
which is independent of $x$, by averaging the likelihood over the prior.
This abstracted, $x$-independent likelihood is labelled the \textbf{evidence}:
\begin{align}
Z
= \prob{d}{h} &= \sum_i L(x_i) \,\pi(x_i)
\label{eqn:searches_evidence_integral_sum}
\\
&= \int\! L(x) \,\mathrm{d}\pi(x),
\label{eqn:searches_evidence_integral_measure}
\end{align}
where Equation~\ref{eqn:searches_evidence_integral_measure} is notation for
integrating the likelihood function over the prior measure~\cite{
billingsley2008probability,
skilling2010foundations
},
can be imagined as simulating from the model $\mathrm{d}\pi$ and
counting events that hit pass the likelihood filter.
The name `evidence' relates to its application for model
comparison~\cite{mackay2003information}.

Suppose that we have this one event with $\met > 100\,\eV[G]$ ---
did it come from the background or signal?%
\footnote{%
Questioning the origin of individual events is unusual in our context in
\atlas, but is done in other research into rare effects such as neutrino
interactions and black hole mergers~\cite{
opera2010event,
icecube2020event,
Abbott2021event
}.%
}
The answer depends on the signal and background cross-sections in our collider,
but larger evidence for signal does influence any conclusion to move in its
favour.

Expressing the evidence $Z$ as an average over the prior highlights how it
depends both on how large that likelihood gets,
and on what fraction of samples hit large likelihood values.
Maximum likelihood approaches,
which compare models by the largest likelihoods their priors support,
can be practical and give useful upper bounds on the evidence.
This histogram example, however, highlights a limitation since both models
support the same maximum likelihood of
$L(\varepsilon + 100\,\eV[G]) = 1$.
These evidence values here differ only by their prior masses in that bin.

@@

Data only enter the \emph{inputs} to Bayes' rule through the likelihood
function, so the likelihood must contain all relevant information from the
data.
The same is true for the evidence for its reuse as a likelihood elsewhere.
To summarize data for scientific communications, we therefore find a
hierarchy of objects with decreasing complexity and increasing abstraction:
\begin{itemize}
\item the data themselves, with instructions for their interpretation,
\item the likelihood function, or functions for varied interpretations, and
\item evidence values for well-defined, predictive models.
\end{itemize}
Each is a compression of the data, which is easier to interpret in more refined
contexts.
Note that summaries of these objects also be useful, such as maxima
of the likelihood function,
or regions in which it exceeds meaningful thresholds.

@@

Many objections to Bayes' rule stem from a fixation on posterior
distributions from unjustified prior assignments.
Classical works by
Bayes~\cite{bayes1763lii} and
Laplace~\cite{laplace1774stigler} began this focus ---
both assume prior densities that are uniform in chosen coordinates, a practice
which can yield practical posterior inferences,
but is not always justified or useful.
But to report scientific data, such practices are fairly criticized as
arbitrary when they influence the reported results.
Prior assignments are arbitrary.
You are welcome to propose any prediction in any circumstances, and the
scientific response is to test how well it predicts our observations of nature.
Our emphasis on the likelihood functions and evidence values that quantify
those predictions avoids the pitfall of seeking uninformative, neutral, or
unbiased priors by embracing priors as active and explicit predictions.
This likelihood perspective is the modern, practical approach to Bayesian data
analysis~\cite{
mackay2003information,
skilling2004nested,
skilling2006nested,
sivia2006data,
skilling2010foundations,
skilling2008rant
}.

Some sources continue to define Bayesian analysis as the computation of
posterior probabilities given data, often without naming the evidence, and
expressing it only as the sum from
Equation~\ref{eqn:searches_evidence_integral_sum}
--- as just a normalizing constant~\cite{
Neyman1937Outline,
gelman1995bayesian,
gelman2008objections,
DAgostini:1994fjx,
DAgostini:2010hil,
cowan1998statistical,
pdg2022ynf
}.
Bayes' rule has two outputs, and both outputs have uses.
Cutting out the evidence hamstrings Bayesian analysis, leaving only the
semi-Bayesian posterior-only approach and its clear flaws.
A bicycle is stable with two wheels, and falls over if one either
is removed;
Figure~\ref{fig:searches_baton_roue_bayes} illustrates this analogy.
A unicycle is more of a circus trick than a practical transport.

\subsection{Probability assignments}
Probability algebra manipulates probabilities, but does not state how they
should be initially assigned.
Such assignments are must be normalized, but might be loosely informed
by calibration data, established theories like quantum mechanics,
or vaguely motivated opinions,
But there are examples of unambiguously derived distributions with common use
in High-energy physics: the Poisson and Gaussian distributions,
both of which can derive from the identity that
\begin{equation}
\label{eqn:searches_exponential}
\exp(x) =
\lim_{N \to \infty}\!
\left(1 + \frac{x}{N}\right)^N
.
\end{equation}
As discussed in Section~\ref{sec:atlas_trigger}, the trigger rejects most
LHC collision events, so our data have fixed large number $N$ of trials
(bunch crossings) and a small probability of acceptance (triggering).
The number of accepted events  $n$ is therefore described by the
binomial distribution:
\begin{equation}
\label{eqn:searches_binomial}
\mathrm{Binomial}(n\mid N, p) = \binom{N}{n}\, p^n (1 - p)^{N - n}
,
\end{equation}
which is a product of the probabilities (product rule) in the sequence of $N$
accept/reject decisions, summed over all possible permutations (sum rule) of
the accepted events.

@@

The Gaussian form arises from products of many functions, which often arise
from the products of likelihood functions on independent data.
Consider a smooth (likelihood) function $f(x)$.
At its maximum $\check{x}$, its first derivative vanishes and its second
derivative is negative, so to leading order a product of $f$s looks like
\begin{equation}
f(x)^N =
\lim_{N \to \infty}\!
\left(
1 - \kappa (x - \check{x})^2
\right)^N,
\end{equation}
which by Equation~\ref{eqn:searches_exponential} goes into the Gaussian form
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma)
\,\mathrm{d}x =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp\!\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)
\mathrm{d}x
,
\end{equation}
which is normalized by its prefactor, and in which we include the
$\mathrm{d}x$ terms as reminders that it is a density function in $x$.
Since the Gaussian form is an exponentiated quadratic, Gaussian likelihoods
justify many least-squares methods.

This is not the Central Limit Theorem, but can be used to derive it: the
distribution of a sum is a convolution which becomes a product in
Fourier space, and that product converges to a Gaussian function by
if the mean and variance exist.
Another more beautiful proof of the Central Limit Theorem is given by
Jaynes~\cite{jaynes2003probability}
(Section ``7.16 The central limit theorem''), but would be spoiled by a
reproduction here.

@@

\subsection{Decision theory}
Now that we have thoroughly covered probability theory, we are ready to return
to the jelly bean game of Figure~\ref{fig:searches_beans}.
Given all you know about jars and beans, you could make a vague guess without
looking at the figure, and you could refine that guess by inspecting the image.
Although I do not propose that human brains \emph{do} use probability
algebra~\cite{jaynes1988brain}, this inference process could be modelled
numerically by distributing probability over the bean numbers
and updating those probabilities (from prior to posterior) in light of a
likelihood function for the image.

@@

Although alternative ideas (such as minimax, minimizing a maximum
loss~\cite{savage1951review}) can work well in certain contexts,
Wald's theorem states that restricting oneself to minimizing an expected cost
function cannot deny one access to optimal decisions.
Alternative formulations can at best be equivalent.

This theory does not define how to construct that cost function.
Like the prior, it must come from context, and indeed the two are intertwined
in the risk;
decision-making machines (or organisms) do learn to make good
decisions without constructing either cost or prior explicitly.
Towards the goal of usefully reporting data, decision theory has been
applied to argue for ways to summarize posterior distributions.
For example, minimizing the expected absolute error
$c(t, x) = |t - x|$ leads
to reporting the median, and minimizing the expected square error
$c(t, x) = (t - x)^2$ leads to reporting the mean.
Had the rules of our bean counting game rewarded you based on such distances,
these results could have been good guesses.
But those weren't the rules.

@@

We do think of many natural processes as random, so this definition works
widely for likelihood functions on data.
Non-data hypotheses, however, are rarely considered as actually random events,
so the frequency definition cannot apply ---
prior predictions often express subjective uncertainties that are manifestly
non-random.
Under this frequency definition, those predictive probabilities are nonsense
and must be rejected.
The frequency definition does not change probability algebra, it only restricts
it to a permitted scope.

@@

Rosencrantz and Guildenstern, in their play by Tom~Stoppard~\cite{
stoppard1967rosencrantz
}
toss a coin ninety-two times, all landing heads.
They then play a game with a third player and land eight more heads.
Will the next toss be heads or tails?
The answer is not random --- it is burned onto paper in print.
If you were to put probabilities on the outcomes, they would be subjectively
informed by it being the one-hundred-and-first toss, our cultural enjoyment
of round numbers in base-10, and your reading of these words.
Frequentist theory, however, forbids such subjective probabilities of
non-random unknowns, and is left searching for alternative forms of inference.

@@

Some authors name this Fisher/Neyman-Pearson assemblage
``classical'' statistics~\cite{
Neyman1937Outline,
lehmann2011fisher,
Feldman:1997qc
},
but that word is simply inaccurate for 20th century innovations on 18th century
ideas~\cite{
bayes1763lii,
laplace1774stigler
}.
Jaynes instead names them ``orthodox''~\cite{
Jaynes1976intervals,
jaynes2003probability
},
which may have been accurate in his environment, but different
(heretical?) ideas have taken hold in research at the LHC ---
orthodoxy has moved on.
We therefore stick with `frequentist'.
Sections~\ref{sec:searches_fisher} and~\ref{sec:searches_np} now introduce
these frequentist methods to explain the historical context for our modern
methods, which are introduced in Section~\ref{sec:searches_practice}.

@@

Fisher's \pvalue\ ratio of $1000$ can therefore be a $\log$ likelihood
ratio of $+6.9$ in favour of $r=0.3$ over $r=0.95$, which is a moderate
indication.
A likelihood ratio on the test statistic itself, and not the inclusive
tail statement, would however encode more information and could provide a
more precise answer.

@@

How can one escape this trap?
Likelihood ratios on the actual data escape immediately by bypassing the
censor, but even inequalities need not be biased --- if taken against a
predetermined threshold, say $5$ events, then models can predict whether
$n \leq 5$, and the honestly answered truth or falsehood in data can favour
either side.
Ultimately, a reader may choose to distrust the reporter ---
the messages
`logically, $n \leq 5$' and
`\atlas\ reports $n \leq 5$' and  are not
identical, since one might infer from the latter that in fact the original
data were $n = 5$, before their reduction into a test statistic tail.
The message itself is data, and what it says may not be the whole truth.

>>>>

\begin{singlespacing}
\chapter{Data analysis and \atlas\ searches}
\label{chapter:searches}
\begin{epigraphs}
\qitem{%
One way is to make it so simple that there are \emph{obviously} no deficiencies
and the\\
other way is to make it so complicated that there are no \emph{obvious}
deficiencies.%
}%
{Tony~Hoare,
\textit{The Emperor’s Old Clothes},
1981~\cite{hoare2007emperor}}
\end{epigraphs}
\end{singlespacing}
\noindent
Our aim in scientific data analysis is to distil useful messages from the
data, and to report that message to the wider community.
Various messages could be extracted from any given data,
and those messages will have varied values to their recipients.
Data analysis is not an automatic process, but a design with many judgements
that are informed by theory, intuition, and convention to give value to their
results.

To support our later description of the \atlas\ SUSY `$\twoljets$' search in
Chapter~\ref{chapter:2ljets}, this chapter introduces the basic theory of
data analysis and its conventional form in the \atlas\ SUSY
community, parts of which can only be explained in their historical context.

@@

\section{Data analysis}
\label{sec:searches_data_analysis}
How many jelly beans are in the jar of Figure~\ref{fig:searches_beans}?
Your guess depends not only on what you observe from the figure, but also on
what else you know about cartoon jars of beans, and on what reward is offered.
Each answer is a decision, which depends both on how you value the rewards,
and on which numbers of beans you subjectively infer to be probable.
We return to this example after developing its analysis in probability
and decision theory.

@@

Probabilistic data analysis follows the ordinary structure of science:
write down predictive models, test their predictions against data,
then update your models (or invent new ones) to refine their predictions.
Likelihood functions are particularly important to this process because
they quantify predictions.
But not everything is a likelihood.

We begin with what probability is: at its most abstract, a probability is a
proportion of stuff within a larger whole.
Probabilities are useful because proportions have important applications,
and because they can be manipulated and interrelated with an intuitive linear
algebra.
The remainder of this section uses the omnipresent tool of high-energy
physics --- the histogram --- to introduce the structure of probability and
its technical language.

\subsection{Probability: abstract}
A probability $\prob{b}{a}$ is a number which quantifies the proportion
of $b$ within $a$, and can be expressed as a ratio
\begin{equation}
\label{eqn:searches_prob_ratio}
\prob{b}{a} = \frac{m(b, a)}{m(a)},
\end{equation}
in which $m(x)$ is a measure, and the comma `,' means `and'
(logical `and', or set intersection or any other equivalent
thing)~\cite{axioms1010038}.

@@

Logic is an important application of probability.
To apply to logical reasoning, $a$ and $b$ are logical propositions and
$\prob{b}{a}$ quantifies the degree to which $a$ implies $b$, with the
limiting cases that $\prob{b}{a} = 1$ when $a$ implies $b$,
and $\prob{b}{a} = 0$ when $a$ contradicts $b$.
Indeed Cox's theorem shows that probability algebra is uniquely consistent
with Boolean logic~\cite{
cox1946probability,
cox1961algebra,
garrett1998nand,
jaynes2003probability,
keynes1920treatise
}.
Probability works for logic, and it works \emph{for tea, too}.

@@

Probability algebra derives from
Equation~\ref{eqn:searches_prob_ratio}; it has a \textbf{product rule}
\begin{align}
\prob{c, b}{a} &= \prob{c}{b, a}\times \prob{b}{a},
\label{eqn:searches_product_rule}
\intertext{%
for nested proportions
(a spoon in a teacup in a sink),
and a \textbf{sum rule}%
}
\prob{c\;\textrm{or}\;b}{a} &= \prob{c}{a} + \prob{b}{a} - \prob{c, b}{a},
\label{eqn:searches_sum_rule}
\end{align}
for combinations in a union, with sums over exhaustive sets of outcomes
necessarily normalized to unity.

@@

Since prior probabilities are normalized distributions, it is usually
convenient to represent them as density functions.
Densities are stated relative to coordinates (or measures),
so changing coordinates changes their values as their unit probability mass
is sloshed around.
Likelihoods are not densities,
so changing coordinates instead slides them horizontally like beads on an
abacus, and does not change their values for equivalent states.
Figure~\ref{fig:searches_abacus} illustrates this difference.

@@

To illustrate this, we consider a histogram that bins event yields in
$\met$, with distributions from background and (supersymmetric) signal samples
overlaid.
Example such histograms are displayed in
Figure~\ref{fig:searches_sig_bkg_prior_likelihood}.
Since these histograms are normalized to unit area, they show prior
probability distributions along the bin axis.
There is one prior for events from each of the background and signal
hypotheses ---
priors distribute unit probability explore along the bin axis.
In each bin, however, a likelihood function compares the assignments from
the two hypotheses: signal and background ---
likelihoods explore vertically down the sample axis.

@@

Both distributions are equivalently drawn in two different coordinates
on the left and right of Figure~\ref{fig:searches_sig_bkg_prior_likelihood};
they are identical in content, but differ in appearance because they represent
each prior distribution as a density in the standard way.%
\footnote{%
Density is a somewhat misleading term here ---
probability mass is incompressible; it moves and redistributes more like
depth of water than density of air.%
}
To preserve the area of each bin, which shows its probability mass,
bin heights change to adapt to their width in each coordinate.
Coordinate-dependence of the prior density means that its point of
maximum density is not always meaningful.
But since the probability mass in each bin is constant, the likelihood is
coordinate-independent.

Standard methods approximate such histograms from Monte Carlo simulations.
Simulations do not directly indicate bin indices, but state the values for
event variables;
to connect event variables to bin indices, we begin with some notation:
we have an hypothesis label $h$ (signal or background),
an event variable $x$ that we histogram ($x = \met$),
and we observe a datum $d$ that indexes the histogram bins.
Supposing, for example, that an event is observed in the overflow bin, it says
$d = `x > 100\,\eV[G]\textrm{'}$.
This observed bin has a likelihood function
\begin{equation}
L(x) = \prob{d}{x, h} = \prob{d}{x} =
\left\{
\begin{matrix}
1 & \textrm{if }~x > 100\,\eV[G], \\
0 & \textrm{otherwise} \\
\end{matrix}
\right.
,
\end{equation}
which corresponds to Equation~\ref{eqn:searches_likelihood} by
$b \leftarrow d$ and $a \leftarrow `x, h\textrm{'}$,
and simplifies by $\prob{d}{x, h} = \prob{d}{x}$ because at given $x$ the
binning is independent of any origin from signal or background.%
\footnote{%
Independence of the likelihood function from prior hypotheses is so common
that it should usually be assumed unless otherwise stated.
}
This example of a binary selection highlights how the likelihood acts as a
filter that only rejects events that are contradicted by the data.
Other likelihoods that are not binary act as inefficient filters
which reject events with some probability.

@@

When we fill a histogram by summing weights of simulated samples, we are
calculating another likelihood over hypotheses $h$ alone in the bin,
which is independent of $x$, by averaging the likelihood over the prior.
This abstracted, $x$-independent likelihood is labelled the \textbf{evidence}:
\begin{align}
Z
= \prob{d}{h} &= \sum_i L(x_i) \,\pi(x_i)
\label{eqn:searches_evidence_integral_sum}
\\
&= \int\! L(x) \,\mathrm{d}\pi(x),
\label{eqn:searches_evidence_integral_measure}
\end{align}
where Equation~\ref{eqn:searches_evidence_integral_measure} is notation for
integrating the likelihood function over the prior measure~\cite{
billingsley2008probability,
skilling2010foundations
},
can be imagined as simulating from the model $\mathrm{d}\pi$ and
counting events that hit pass the likelihood filter.
The name `evidence' relates to its application for model
comparison~\cite{
mackay2003information,
skilling2004nested,
skilling2006nested
}.

Suppose that we have one event with $\met > 100\,\eV[G]$.
Did it come from the background or signal?%
\footnote{%
Questioning the origin of individual events is unusual in our context in
\atlas, but is done in other research into rare effects such as neutrino
interactions and black hole mergers~\cite{
opera2010event,
icecube2020event,
Abbott2021event
}.%
}
The answer depends on the signal and background cross-sections in our collider,
but larger evidence for signal does influence any conclusion to move in its
favour.

Expressing the evidence $Z$ as an average over the prior highlights how it
depends both on how large that likelihood gets,
and on what fraction of samples hit large likelihood values.
Maximum likelihood approaches,
which compare models by the largest likelihoods their priors support,
can be practical and give useful upper bounds on the evidence.
This histogram example, however, highlights a limitation since both models
support the same maximum likelihood of
$L(101\,\eV[G]) = 1$.
These evidence values differ only by their prior masses in the bin.

@@

Data only enter the \emph{inputs} to Bayes' rule through the likelihood
function, so the likelihood must contain all relevant information from the
data; the same is true for the evidence for its reuse elsewhere as an
abstracted likelihood.
To summarize data for scientific communications, we therefore find a
hierarchy of objects with decreasing complexity and increasing abstraction:
\begin{itemize}
\item the data themselves with instructions for their interpretation,
\item the likelihood function, or functions for varied interpretations, and
\item evidence values for well-defined, predictive models.
\end{itemize}
Each step compresses the data, into a smaller form that is easier to interpret
in more refined contexts.
Summaries of these objects are useful, such as maxima of the likelihood function,
or regions in which it exceeds meaningful thresholds.

@@

Common objections to Bayes' rule stem from a fixation on posterior
distributions from unjustified prior assignments.
Classical works by
Bayes~\cite{bayes1763lii} and
Laplace~\cite{laplace1774stigler} began this focus ---
both assume prior densities that are uniform in chosen coordinates, a practice
which can yield practical posterior inferences,
but is not always justified or useful.
But to report scientific data, such practices are fairly criticized as
arbitrary when they influence the reported results.
Prior assignments are arbitrary.
Our emphasis on the likelihood functions and evidence values that quantify
those predictions avoids the pitfall of seeking uninformative, neutral, or
unbiased priors by embracing priors as active and explicit predictions.
This likelihood perspective is the modern, practical approach to Bayesian data
analysis~\cite{
mackay2003information,
skilling2004nested,
skilling2006nested,
sivia2006data,
skilling2010foundations,
skilling2008rant
}.

Some sources continue to define Bayesian analysis as the computation of
posterior probabilities given data, often without naming the evidence, and
expressing it only as the sum from
Equation~\ref{eqn:searches_evidence_integral_sum}
--- as just a normalizing constant~\cite{
Neyman1937Outline,
gelman1995bayesian,
gelman2008objections,
DAgostini:1994fjx,
DAgostini:2010hil,
cowan1998statistical,
pdg2022ynf
}.
Bayes' rule has two outputs, and both outputs have uses.
Cutting out the evidence hamstrings Bayesian analysis, leaving only the
semi-Bayesian posterior-only approach and its clear flaws.
A bicycle is stable with two wheels, and falls over if one either
is removed;
Figure~\ref{fig:searches_baton_roue_bayes} illustrates this analogy;
a unicycle is more of a circus trick than a practical transport.

\subsection{Probability assignments}
Probability algebra manipulates probabilities, but does not state how they
should be initially assigned.
Such assignments are must be normalized, but might be loosely informed
by calibration data, established theories like quantum mechanics,
or vaguely motivated opinions,
But there are examples of unambiguously derived distributions with common use
in high-energy physics: the Poisson and Gaussian distributions,
both of which can derive from the identity that
\begin{equation}
\label{eqn:searches_exponential}
\exp(x) =
\lim_{N \to \infty}\!
\left(1 + \frac{x}{N}\right)^N
.
\end{equation}
As discussed in Section~\ref{sec:atlas_trigger}, the trigger rejects most
LHC collision events, so our data have fixed large number $N$ of trials
(bunch crossings) and a small probability of acceptance (triggering).
The number of accepted events  $n$ is therefore described by the
binomial distribution:
\begin{equation}
\label{eqn:searches_binomial}
\mathrm{Binomial}(n\mid N, p) = \binom{N}{n}\, p^n (1 - p)^{N - n}
,
\end{equation}
which is a product of the probabilities (product rule) in the sequence of $N$
accept/reject decisions, summed over all possible permutations (sum rule) of
the accepted events.

@@

The Gaussian form arises from products of many functions, which often arise
from the products of likelihood functions on independent data.
Consider a smooth (likelihood) function $f(x)$.
At its maximum $\check{x}$, its first derivative vanishes and its second
derivative is negative, so at leading order a product of $f$s near its maximum
looks like
\begin{equation}
f(x)^N =
\lim_{N \to \infty}\!
\left(
1 - \kappa (x - \check{x})^2
\right)^N,
\end{equation}
which by Equation~\ref{eqn:searches_exponential} goes into the Gaussian form
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma)
\,\mathrm{d}x =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp\!\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)
\mathrm{d}x
,
\end{equation}
which is normalized by its prefactor, and in which we include the
$\mathrm{d}x$ terms as reminders that it is a density function in $x$.
Since the Gaussian form is an exponentiated quadratic, Gaussian likelihoods
justify many least-squares methods.

This is not the famous central limit theorem, but can be used to derive it: the
distribution of a sum is a convolution which becomes a product in
Fourier space, and that product converges to a Gaussian function by
if the mean and variance exist.
A more beautiful proof of the central limit theorem is given by
Jaynes~\cite{jaynes2003probability}
(Section ``7.16 The central limit theorem''), but would be spoiled by
reproduction here.

@@

\subsection{Decision theory}
Now that we have thoroughly covered probability theory, we  return
to the jelly bean game of Figure~\ref{fig:searches_beans}.
Given all you know about jars and beans, you could make a vague guess without
looking at the figure, and you could refine that guess by inspecting the image.
Although I do not propose that human brains \emph{do} use probability
algebra~\cite{jaynes1988brain}, this inference process could be modelled
numerically by distributing probability over the bean numbers
and updating those probabilities (from prior to posterior) in light of a
likelihood function for the image.

@@

Although alternative ideas (such as minimax, minimizing a maximum
loss~\cite{savage1951review}) can work well in certain contexts,
Wald's theorem states that restricting oneself to minimizing an expected cost
function cannot deny access to optimal decisions.
Alternative formulations can at best be equivalent.

This theory does not define how to construct a cost function.
Like probability assignments, it must come from context, and indeed the two
are intertwined in the risk;
decision-making machines (or organisms) do learn to make good
decisions without constructing either cost or prior explicitly.
Towards the goal of usefully reporting data, decision theory has been
applied to argue for ways to summarize posterior distributions.
For example, minimizing the expected absolute error
$c(t, x) = |t - x|$ leads
to reporting the median, and minimizing the expected square error
$c(t, x) = (t - x)^2$ leads to reporting the mean.
Had the rules of our bean counting game rewarded you based on such distances,
these results could have been good guesses.
But those weren't the rules.

@@

We do think of many natural processes as random, so this definition works
widely for likelihood functions on data.
Non-data hypotheses, however, are rarely considered as actually random events,
so the frequency definition cannot apply ---
prior predictions often express subjective uncertainties that are manifestly
non-random.
Under this frequency definition, those predictive probabilities are nonsense
and must be rejected.
The frequency definition does not change probability algebra,
but it does restrict its scope.

@@

Rosencrantz and Guildenstern, in their play by Tom~Stoppard~\cite{
stoppard1967rosencrantz
}
toss a coin ninety-two times, all landing heads.
They then play a game and land eight more heads.
Will the next toss land heads or tails?
The answer is not random --- it is burned onto paper in print.
If you were to put probabilities on the outcomes, they would be subjectively
informed by it being the one-hundred-and-first toss, our cultural enjoyment
of round numbers in base-10, and your reading of these words.
Frequentist theory forbids such subjective probabilities of
non-random unknowns, and is left searching for alternative forms of inference.

@@

Some authors name this Fisher/Neyman-Pearson assemblage
``classical'' statistics~\cite{
Neyman1937Outline,
lehmann2011fisher,
Feldman:1997qc
},
but that word is simply inaccurate for 20th century innovations on 18th century
ideas~\cite{
bayes1763lii,
laplace1774stigler
}.
Jaynes instead names them ``orthodox''~\cite{
Jaynes1976intervals,
jaynes2003probability
},
which may have been accurate in his environment, but different
(heretical?) ideas have taken hold in research at the LHC ---
orthodoxy has moved on.
We therefore stick with `frequentist'.
Sections~\ref{sec:searches_fisher} and~\ref{sec:searches_np} introduce
these frequentist methods to explain the historical context of our modern
methods, which are introduced in Section~\ref{sec:searches_practice}.

@@

Fisher's \pvalue\ ratio of $1000$ can therefore be a $\log$ likelihood
ratio of $+6.9$ in favour of $r=0.3$ over $r=0.95$, which is a notable
indication.
A likelihood ratio on the test statistic or data itself would however
encode more information and could provide a more precise answer.

@@

How can one escape this trap?
Likelihood ratios on the actual data escape immediately by bypassing the
censor, but even inequalities need not be biased if models can ask
questions of the data and be given honest answers.
For example, if asking whether the data exceed a predetermined threshold
of  $5$ events, then the truth or falsehood of `$n \leq 5$'
can favour either side.
